---
title: "Linguistic Data Analysis - Final Project"
author: "Benedict Wuethrich"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#register_stadiamaps(key = "ENTER YOUR OWN KEY HERE",
#                    write = FALSE)
```


Everything related to this project can be found in [this GitHub repository](https://github.com/BendixY/LDAR-FP-BW).

```{r packages, include=FALSE}

library(tidyverse)
library(here)
library(ggmap)
library(ggthemes)
library(maps)
library(viridis)
```
## Introduction
On the Australian continent there are over 333 reported languages which can be roughly categorized into either Pama-Nyungan languages or Non-Pama-Nyungan languages. It is important to note that the latter does not imply any genealogical link of the included languages, while the former has been shown as a cohesive language family. Non-Pama-Nyungan is more of a collective term for Australian languages that aren't thought of as Pama-Nyungan. Exactly which langauges belong to which category is an ongoing debate.

Pama-Nyungen languages have been spoken for over 5000 years, make up over 306 different identified languages and the speakers cover about 80% of the landmass (Bouckaert 2018, 741). Given this long history, it is likely that the Pama-Nyungan languages have been competing against Non-Pama-Nyungan languages for a long time and ended up limiting the Non-Pama-Nyungan languages to the northernmost part of the continent. Nowadays, both are having to compete against the English language.


## Data Wrangling
The data wrangling for this project was relatively simple, since the datasets provided by WALS and phoible respectively are quite clean already. 
```{r data_wrangling, include=FALSE}
#loading in data from phoible and WALS and limiting them to languages spoken in Australia

phoible <- read_csv(here("data/phoible/phoible.csv")) 
glimpse(phoible)

wals_aus <- read_csv(here("data/WALS/cldf/languages.csv")) %>% 
  filter( #Latitude >= -44 & Latitude <= -10 & 
           #Longitude >= 112 & Longitude <= 154 &
           Macroarea == "Australia")
#boundaries were determined with the help of ChatGPT 4.o but ended up not being helpful enough. i left the code in in case i might need it at another time
glimpse(wals_aus)
```

```{r AUS_map}

map_AUS <- get_stadiamap(bbox = c(left = 112, 
                                  bottom = -44,
                                  right = 154,
                                  top = -10), #here the map boundaries given by ChatGPT earlier came in handy again
                         zoom = 5,
                         maptype = "stamen_toner_lite",
                         color = "color")
ggmap(map_AUS)



map_AUS_family <- ggmap(map_AUS) +
  geom_point(data = wals_aus,
             aes(x = Longitude,
                 y = Latitude,
                 color = Family),
             size = 9) +
  scale_color_colorblind() + #better color
  theme_map() + #removes axes labels and puts the legend in bottom left corner of map
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size =12)) +
  #        legend.position.inside = c(0.9,0)) + #we can still modify it to our liking 
  labs(color = "Families")
map_AUS_family

```


## Bibliography

Bouckaert, Remco R., Claire Bowern & Quentin D. Atkinson (2018). The origin and expansion of Pama–Nyungan languages across Australia. Nature Ecology & Evolution (2), 741–749.  
Bowern, Claire / Koch, Harold (Hrsg.) (2004): Australian Languages. Classification and the comparative method. Amsterdam / Philadelphia: John Benjamins.  
Dixon, R.M.W. (1980): The Languages of Australia. Cambridge: Cambridge University Press.  
Dixon, R.M.W. (2004): Australian Languages. Their Nature and Development. Cambridge: Cambridge University Press.  
Kahle, D., Wickham, H. ggmap: Spatial Visualization with ggplot2. The R Journal, 5(1), 144-161. http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf  
Matthew S. Dryer. (2013) Order of Subject, Object and Verb. In: Dryer, Matthew S. & Haspelmath, Martin (eds.) WALS Online (v2020.3) [Data set]. Zenodo. https://doi.org/10.5281/zenodo.7385533 (Available online at http://wals.info/chapter/81, Accessed on 2024-04-16.)  
O’Grady, G. N. (1998). Toward a Proto-Pama-Nyungan Stem List, Part I: Sets J1-J25. Oceanic Linguistics, 37(2), 209–233.  
Schmidt, W. (1919). Die Gliederung der australischen Sprachen: geographische, bibliographische, linguistische Grundzüge der Erforschung der australischen Sprachen. Mechitharisten-Buchdruckerei.  
Wickham, H et al. (2019). Welcome to the tidyverse. Journal of Open Source Software,  4 (43), 1686. doi:10.21105/joss.01686 https://doi.org/10.21105/joss.01686  
Zuckermann, G. et al. (2021) LARA in the Service of Revivalistics and Documentary Linguistics: Community Engagement and Endangered Languages. Proceedings of the Workshop on Computational Methods for Endangered Languages (1), 13-23.  

